Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-15 Response.tex

\bigskip
\medskip
\noindent{\Large 
Response 
}
\bigskip

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%---SUGGESTED SIMULATION SCENARIOS---%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Simulations}

\begin{itemize}

\item {
{\it
\textbf{
Simulate a number of epochs that do not correspond with the assumed ones.
}% END: textbf
}% END: it
}% END: item
\\
FB: Perhaps two sets of simulations here: \\
* simulate 2 epochs - inference under 1 \\
* simulate 3 - inference for 2 \\
Check the variance and bias

\item {
{\it
\textbf{
Simulating under a non-epoch model, for example where different process can apply to contemporaneous branches, would be fundamental to understand the robustness of the model.
}% END: textbf
}% END: it
}% END: item
\\
FB: I think he means we did simulations with an ultrametric tree, but we simulated and recovered the values under the same (epoch) model.
For that I could set up a scenario where data is generated under e.g. HKY(1.0) with an underlying ultrametric topology and then we recover two epochs, placing the transition time somewhere arbitrary.

\item {
{\it
\textbf{
Here it would have been interesting to see the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model).
}% END: textbf
}% END: it
}% END: item
\\
FB: I think here we could repeat the study from the $\pi$BUSS paper but simulate under a homogenous model. We indeed could set the $\omega$ parameter to a weighted mean of the values we used previously (1.0, 0.2, 0.1, 0.02, 0.01, 0.002, 0.001), as weights we could use epoch length, where for last epoch it's tMRCa - T[n]. At least that's how I understand this comment.

\item {
{\it
\textbf{
How mislead one could be if the data were generated under an epoch model, but analyzed under a simpler model.
}% END: textbf
}% END: it
}% END: item
\\
I think this comment is asking basically the same thing as David Posada's first sugested simulation.

@@

% 3
\item {
{\it
\textbf{
In general, there is some repetitive text, sometimes with identical phrases/paragraphs, between the introduction and some of the other chapters. 
On the other hand, some relevant aspects of the research are discussed in the general discussion in Chapter 5, instead of in the corresponding `research' chapter. 
This is true in particular for Chapter 3, were a very concise description of SPREAD is provided. 
I understand these redundancies and fragmentations are a byproduct of reproducing papers as chapters, but however these issues have been corrected in some occasions. 
For example, chapter 2 is not identical to the SysBio paper. 
I suggest you try a bit to further consolidate the different chapters.
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 4
\item {
{\it
\textbf{
The simulation software piBUSS described in Chapter 4 can be very useful, in particular in regards to its connectivity with BEAST. 
The possibility of simulating discrete phylogeographic traits is very convenient. 
Here it would have been interesting to see, for the sake of comparison and interpretation, the correspondence between simulated and estimated tMRCAs when the true model has one epoch with omega=0.5 (or even better, omega equal to the weighted mean of the n-epoch model). 
This does not need to be done before the defense, but it could be a nice extension for the future.
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 5
\item {
{\it
\textbf{
Page 14: I would say that the term `Big Data' does not apply to sequence alignments. 
It might do for NGS files etc., but not for the type of data that BEAST uses as input. 
Not yet. 
Moreover, you focus on HPC architectures like GPU and multicores, which are in some senses opposite to Big Data architectures (i.e., Hadoop Map Reduce).
}% END: textbf
}% END: it
}% END: item
\\
Rephrased to avoid term:

\begin{quote}
\myeditDPfive
\end{quote}

@@

% 6
\item {
{\it
\textbf{
Page 25, 31: likelihood should be $P(X| \theta)$, not $P(\theta | X)$.
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 7
\item {
{\it
\textbf{
Page 34: on the left-term of equation 1.37, the denominator should be
$P(\theta[t]|X) q(\theta^{*} | \theta[t])$
not 
$P(\theta[t]|X) q(\theta[t] | \theta^{*})$
}% END: textbf
}% END: it
}% END: item
\\
AAA

% 8
\item {
{\it
\textbf{
Page 52: according to the legend in Figure 2.3, a two-epoch plot is missed.
}% END: textbf
}% END: it
}% END: item
\\
Hmm I think it wasn't clear that it's all on one plot. Needs re-phrasing:

% \begin{quote}
% \myeditDPeight
% \end{quote}

### 2015-01-15 Response.tex

\end{itemize}

%%%%%%%%%%%%%%%%%%%%
%--- COMMENTS---%
%%%%%%%%%%%%%%%%%%%%
\section*{Comments by David Posada}

\begin{itemize}

% 1
\item {
{\it
\textbf{
I missed a brief primer of Bayesian model selection in the introductory chapter.
Techniques such Bayes factors, path sampling or stepping-stone sampling are used throughout but never explained. 
This should be very easy to fix.
}% END: textbf
}% END: it
}% END: item
\\
There is now a subsection called \emph{Bayesian model selection} at the end of Chapter 1.

@@

% 6
\item {
{\it
\textbf{
Page 25, 31: likelihood should be $P(X| \theta)$, not $P(\theta | X)$.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 7
\item {
{\it
\textbf{
Page 34: on the left-term of equation 1.37, the denominator should be
$P(\theta[t]|X) q(\theta^{*} | \theta[t])$
not 
$P(\theta[t]|X) q(\theta[t] | \theta^{*})$
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
Page 52: according to the legend in Figure 2.3, a two-epoch plot is missed.
}% END: textbf
}% END: it
}% END: item
\\
Hmm I think it wasn't clear that it's all on one plot. Needs re-phrasing:

\begin{quote}
{\bf  Epoch simulation scenarios on an influenza A maximum clade credibility tree topology.} 
In the two-epoch example illustrated by the top alternating background
% at the top, 
the transition time is set at $t_{1}=7$, creating two epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$ and $\mathbf{Q_{2}}$ respectively, separated by the light and dark grey areas and the dotted line.
In the three-epoch example illustrated 
by the bottom alternating background
% at the bottom, 
transition times are put at $t_{1}=7$ and $t_{2}=15$, creating three epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$, $\mathbf{Q_{2}}$ and then again $\mathbf{Q_{1}}$, as indicated by the alternating dark and light areas and the dotted lines.
\end{quote}

### 2015-01-15 chapters/chapter01.tex
subsection on Bayesian model selection

\begin{equation}
L\left(\mathbf{\Theta}|\mathbf{X}\right)=P\left(\mathbf{\Theta}|\mathbf{X}\right), 
\label{eq:likelihood}
\end{equation}

@@

% TODO: discuss hyperpriors?
The prior information is then combined with the likelihood of the data $P\left(\theta|\mathbf{X}\right)$ to form the posterior, via Bayes' theorem (Equation~(\ref{eq:bayesian1})).
The posterior distribution of $\theta$ follows:

@@

\begin{equation}
P\left(\theta|\mathbf{X}\right) =\int P\left(\mathbf{\Theta},\mathbf{F}|\mathbf{X}\right)d\mathbf{F}
\label{eq:bayesian3}
\end{equation}

@@

\begin{equation}
\frac{P\left(\theta^{*}|\mathbf{X}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P\left(\theta\left[t\right]|\mathbf{X}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}=\frac{P(\theta^{*})\cdot P\left(\mathbf{X}|\theta^{*}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P(\theta\left[t\right])\cdot P\left(\mathbf{X}|\theta\left[t\right]\right)\cdot q\left(\theta^{*}|\theta\left[t\right]\right)}.
\label{eq:metropolis1}
\end{equation}

@@

>>>>

\begin{equation}
L\left(\mathbf{\Theta}|\mathbf{X}\right)=P\left( \mathbf{X} | \mathbf{\Theta} \right), 
\label{eq:likelihood}
\end{equation}

@@

% TODO: discuss hyperpriors?
The prior information is then combined with the likelihood of the data $P\left( \mathbf{X} | \theta \right)$ to form the posterior, via Bayes' theorem (Equation~(\ref{eq:bayesian1})).
The posterior distribution of $\theta$ follows:

@@

\begin{equation}
P\left(\theta|\mathbf{X}\right) =\int P\left(\theta,\mathbf{F}|\mathbf{X}\right)d\mathbf{F}
\label{eq:bayesian3}
\end{equation}

@@

% \myedit{DPseven}{
\begin{equation}
\frac{P\left(\theta^{*}|\mathbf{X}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P\left(\theta\left[t\right]|\mathbf{X}\right)\cdot q\left(\theta^{*}|\theta\left[t\right]\right)}=\frac{P(\theta^{*})\cdot P\left(\mathbf{X}|\theta^{*}\right)\cdot q\left(\theta\left[t\right]|\theta^{*}\right)}{P(\theta\left[t\right])\cdot P\left(\mathbf{X}|\theta\left[t\right]\right)\cdot q\left(\theta^{*}|\theta\left[t\right]\right)}.
\label{eq:metropolis1}
\end{equation}
% }

@@

\subsection{Bayesian model selection\label{sub:model_selection}}
% https://radfordneal.wordpress.com/2008/08/17/the-harmonic-mean-of-the-likelihood-worst-monte-carlo-method-ever/

A standard approach to compare models in a Bayesian phylogenetic framework is through the evaluation of Bayes factors (BFs).   
The BF is a ratio of two marginal likelihoods, which given the observed data $\mathbf{X}$ quantify the plausibility of two models, denoted $M_{1}$ and $M_{2}$ and parametrised in terms of parameters $\theta_{1}$ and $\theta_{2}$ respectively:

% 
\begin{equation}  
\frac{\int P\left(\mathbf{X}|\theta_{1},M_{1}\right)\times P\left(\theta_{1} \mid M_{1}  \right)d_{\theta_{1}}}{\int P\left(\mathbf{X}|\theta_{1},M_{2}\right)\times P\left(\theta_{2} \mid M_{2} \right)d_{\theta_{2}}},
\label{eq:bayes_factor}
\end{equation}  
%

\noindent
where $P\left(\mathbf{X}|\theta_{i},M_{i}\right)$ is the likelihood function and $P\left(\theta_{i} \mid M_{i} \right)$ is the prior, for $i=1, \ldots, 2$.
The parameters $\theta_{i},\; i=1, \ldots, 2$ can be possibly multidimensional, making the integral in Equation~\ref{eq:bayes_factor} non-trivial to estimate.
For a number of years the posterior Harmonic Mean Estimator (HME, \cite{Newton1994}) has been extesively used in Bayesian phylogenetic inference, mainly for the simplicity in it's calculation. 
To evaluate the HME one only needs the samples from the posterior distribution which are readilly availiable as output form any MCMC program.

\cite{Meng1996} as well as \cite{Xie2011} warn that despite this computational convenience the HME is a biased estimator of the marginal likelihood, severely overestimating and with potential infinite variance.
Realizing this has lead to a substantial amount of work invested in the development of efficient methods for estimating marginal likelihoods \citep{Suchard2001,Lemey2009,Li2012}.
Recently \cite{Lartillot2006} as well as \cite{Xie2011} proposed to incorporate the path sampling (PS) and stepping-stone sampling (SS) methods for accurate estimation of the marginal likelihood into Bayesian phylogenetic frameworks.
\cite{Baele2012} have proposed the first implementation of PS and SS into phylogenetic inference software.
Both these techniques rely on drawing MCMC samples from a series of distributions, each of which is a power posterior differing only in its power, along the path going from the prior to the unnormalized posterior defined by the model $M$:

\begin{equation}  
q_{\beta}\left(\theta\right)=P\left(\mathbf{X}\mid\theta,M\right)\times P\left(\theta\mid M\right).
\label{eq:ss_ps}
\end{equation}

Although offering tremendous improvements in bias and variance over the HME both PS and SS require considerable computational effort as the marginal likelihood can no longer be estimated from the likelihood trace that is being output from the standard MCMC analysis. 
% When only modest computation is invested to keep marginal likelihood manageable for large data sets, the estimates may suffer from considerable variance, preventing an accurate comparison of two models.
Further, great care must also be taken to specify proper priors for all parameters in Equation~\ref{eq:ss_ps}, so as to avoid numerical instabilities when the integration approaches the prior.

### 2015-01-15 chapters/chapter02.tex
proposed simulations
subsection on Bayesian model selection

%---TOPOLOGY FOR SIMULATIONS---%
\begin{figure}[H]
\centering
% \color{fgcolor}
\includegraphics[scale=0.35]{treeplot} 
\caption{
{ \footnotesize 
% \myedit{DPsix}{
{\bf  Epoch simulation scenarios on an influenza A maximum clade credibility tree topology.} 
In the two-epoch example illustrated by the top alternating background
% at the top, 
the transition time is set at $t_{1}=7$, creating two epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$ and $\mathbf{Q_{2}}$ respectively, separated by the light and dark grey areas and the dotted line.
In the three-epoch example illustrated 
by the bottom alternating background
% at the bottom, 
transition times are put at $t_{1}=7$ and $t_{2}=15$, creating three epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$, $\mathbf{Q_{2}}$ and then again $\mathbf{Q_{1}}$, as indicated by the alternating dark and light areas and the dotted lines.
% }%END: myedit
}% END: footnotesize
}%END: caption
\label{fig:2RateTree}
\end{figure}

>>>>

%---TOPOLOGY FOR SIMULATIONS---%
\begin{figure}[H]
\centering
% \color{fgcolor}
\includegraphics[scale=0.35]{treeplot} 
\caption{
{ \footnotesize 
{\bf  Epoch simulation scenarios on an influenza A maximum clade credibility tree topology.} 
% \myedit{DPeight}{
In the two-epoch example illustrated by the top alternating background
% at the top, 
the transition time is set at $t_{1}=7$, creating two epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$ and $\mathbf{Q_{2}}$ respectively, separated by the light and dark grey areas and the dotted line.
In the three-epoch example illustrated 
by the bottom alternating background
% at the bottom, 
transition times are put at $t_{1}=7$ and $t_{2}=15$, creating three epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$, $\mathbf{Q_{2}}$ and then again $\mathbf{Q_{1}}$, as indicated by the alternating dark and light areas and the dotted lines.
% }%END: myedit
}% END: footnotesize
}%END: caption
\label{fig:2RateTree}
\end{figure}

### 2015-01-15 chapters/chapter06.tex
subsection on Bayesian model selection

\noindent
From Equations~(\ref{eq:rodrigo}) and (\ref{eq:approx}), by the definition of the Riemannian integral we have that $\mathbf{P}(r,T)=\underset{n\rightarrow\infty}{lim}\underset{i=0}{\overset{n}{\prod}}\text{exp}\left(r\mathbf{Q}(t_{i}^{*})\cdot\triangle t_{i}\right)$, given that rate matrices $\mathbf{Q}(t_{i}^{*})$ commute 
and are closed with respect to the matrix multiplication.
\citet{Sumner2012} formalize the problem of multiplicative closure of the Markov models. 
Given those regularity conditions, we can perform a numerical approximation of any function of rate change by using the epoch time-discretization, and by partitioning the time interval into a fine grid of intervals.
Extending our Bayesian epoch model implementation to accommodate complex rate change functions could be the focus of future work when confronted with a problem that would benefit from such an approach.

>>>>

\noindent
From Equations~(\ref{eq:rodrigo}) and (\ref{eq:approx}), by the definition of the Riemann integral we have that $\mathbf{P}(r,T)=\underset{n\rightarrow\infty}{lim}\underset{i=0}{\overset{n}{\prod}}\text{exp}\left(r\mathbf{Q}(t_{i}^{*})\cdot\triangle t_{i}\right)$, given that rate matrices $\mathbf{Q}(t_{i}^{*})$ commute 
and are closed with respect to the matrix multiplication.
\citet{Sumner2012} formalize the problem of multiplicative closure of the Markov models. 
Given those regularity conditions, we can perform a numerical approximation of any function of rate change by using the epoch time-discretization, and by partitioning the time interval into a fine grid of intervals.
Extending our Bayesian epoch model implementation to accommodate complex rate change functions could be the focus of future work when confronted with a problem that would benefit from such an approach.
